---
title: "Arbeidskrav 4"
author: "Jonas A."
format: html
editor: visual
warnings: false
---

```{r}
#cleaning enviroment and loading poackages
rm(list = ls())
library(tidyverse)
library(quantmod)
library(lubridate)
```

# Task 1

The table below shows (table 1) the total sales by month and customer segment for the regions 1 and 3.

```{r}
#part 1

#downloading dataset
df <- read.csv("https://raw.githubusercontent.com/uit-sok-1005-v23/uit-sok-1005-v23.github.io/main/storedata.csv")

#makes order_date in date format
df$Order_Date <- as.Date(df$Order_Date)

#extracts year, month and day til new columns for easier filtering
df$Order_Year <- format(df$Order_Date, "%Y")
df$Order_Month <- format(df$Order_Date, "%m")
df$Order_Day <- format(df$Order_Date, "%d")

#makes a duplicate
table1 <- df

#filtering for 2017 and the wanted months, and selecting wanted data
table1 <- table1 %>% 
  filter(Order_Year == "2017", Order_Month >= "10", Customer_Segment %in% c("Corporate", "Consumer"), Region %in% c("Region 1", "Region 3")) %>% #filtering data
  group_by(Region, Order_Month, Customer_Segment) %>% #grouping by wanted args and 
  summarize(Total_Sales = sum(Sales)) # summarizing

print(table1)
```

Figure 1 shows the total sales for region 1 and 3 by year and month.

```{r}
#part 2

figure1 <- df %>% filter(Order_Year %in% c("2015", "2016", "2017"), Region %in% c("Region 1", "Region 13")) %>% 
  group_by(Region, Order_Year, Order_Month) %>%
  summarize(Total_Sales = sum(Sales)) %>%
  ggplot(aes(x = Order_Month, y = Total_Sales, fill = Order_Year)) +
  geom_col(width = 0.5, position = "dodge") +
  facet_wrap(~ Region, ncol = 2) +
  labs(x = "Month", y = "Total Sales", fill = "Year")+
  scale_x_discrete(guide = guide_axis(n.dodge=2)) +
  ggtitle("Figure 1: Total sales by year, month and region") +
  theme_minimal()

figure1
```

In the months shown in the table below (table 2), the total sales were greater in region 13 than in region 1.

```{r}
#part 3

table2 <- df %>% filter(Order_Year %in% c("2015", "2016", "2017"), Region %in% c("Region 1", "Region 13")) %>% 
  group_by(Region, Order_Year, Order_Month) %>%
  summarize(Total_Sales = sum(Sales)) #selecting same years and regions as in fig 1, and also creating the total sales var

table2 <- table2 %>% 
  pivot_wider(names_from = Region, values_from = Total_Sales) %>% #pivot wider to get the regions seperated
  filter(!!sym("Region 1") < !!sym("Region 13")) %>% 
  select(Order_Year, Order_Month, !!sym("Region 1"), !!sym("Region 13")) #filterering where the total sales value in region 13 is greater than in region 1

print(table2)
```

The table below (table 3) shows the which product and consumer segment that made the most profit on average in 2017. We can see that technology sold to small buisnesses made the most average profit.

```{r}
#part 4

table3 <- df %>% #filtering
  filter(Order_Year == 2017, !Region %in% c("Region 3", "Region 5", "Region 8")) %>%
  group_by(Customer_Segment, Product_Category) %>% 
  summarize(Average_Profit = mean(Profit)) %>% #calculating mean
  arrange(desc(Average_Profit)) #descending order

print(table3)
```

# Task 2

The figure below shows the monthly average stock price for exxon, aswell as the monthly average price for oil barrels in dollars. We see some correlation between the stock price of the oil company Exxon Mobile and the oil price.

```{r}
getSymbols("XOM", src = "yahoo") #Data XOM stock price
xom_df <- as.data.frame(XOM)
xom_df <- tibble::rownames_to_column(xom_df, var = "Date")
xom_df$Date <- as.Date(xom_df$Date) #as date
xom_df <- xom_df %>% 
  filter(Date >= "2010-01-04" & Date <= "2022-12-31")

exxon <- xom_df %>%  #making monthly volume weighted mean of the close prices
  mutate(year_month = format(Date, "%Y-%m")) %>% 
  group_by(year_month) %>% 
  summarize(exxon_avg = weighted.mean(XOM.Close, XOM.Volume))

exxon$year_month <- ym(exxon$year_month) #making new year_month var into date var

getSymbols("DCOILBRENTEU", src = "FRED") #Data Brent Crude Oil
oil_df <- as.data.frame(DCOILBRENTEU)
oil_df <- tibble::rownames_to_column(oil_df, var = "Date")
oil_df$Date <- as.Date(oil_df$Date)
oil_df <- oil_df %>%
  filter(Date >= "2010-01-04" & Date <= "2022-12-31")

oil <- na.omit(oil_df) %>%  #calculating monthly avg for oil
  mutate(year_month = format(Date, "%Y-%m")) %>% 
  group_by(year_month) %>% 
  summarize(oil_avg = mean(DCOILBRENTEU))

oil$year_month <- ym(oil$year_month) #making new year_month var into date var

colors <- c("Exxon" = "#B1624EFF", "Oil" = "#5CC8D7FF")

ggplot() + 
  geom_line(data = exxon, aes(x = year_month, y = exxon_avg, color="Exxon")) + #exxon line
  geom_line(data = oil, aes(x = year_month, y = oil_avg, color="Oil")) + #oil line
  xlab("Year-Month") +  #x label
  ylab("Price (USD)") + #y label
  ggtitle("Monthly Average of Exxon Mobile 
and Brent Crude Oil Prices") + #title for plot
  theme_minimal() + #theme
  scale_x_date(date_labels = "%Y-%m") + #date labels
  scale_color_manual(values = colors) + #colors from vector above
  labs(colour = NULL) #remove color legend

```

```{r}
#merging the dfs by equal column
exxon_oil <- merge(exxon, oil, by = "year_month")

exxon_oil %>%
ggplot(aes(x=exxon_avg, y=oil_avg)) +
  geom_smooth(method=lm, se = FALSE)+
  xlim(30, 125)+
  ylim(30, 125) +
  geom_abline(slope=1, intercept = 0, color="red") +
  labs(title="Blue line: Correlation between Exxon
and oil price", y="Oil", x="Exxon") +
  theme_minimal()
```

If the correlation were perfect, the blue line would be the same as the red one. The red line represents a perfect correlation.

```{r}
lm(oil_avg ~ exxon_avg, data=exxon_oil)
```

The code above shows the correlation between the oil price and exxon stock prices. It says the correlation is 0.8492, where 1 is a perfect correlation. We can see the exxon stock and oil price is quite correlated.
